---
title: "Ghana Scenario Planning Notes"
output: html_notebook
---


### Poverty Profile Report 2005 - 2017 

##### GDP Growth and Inflation
* Average annual growth was 6.8 percent  
* In 2017, Ghana was one of the fastest growing economies in the world
* Weaker fiscal and monetary policies, terms of trade shocks from lower oil prices, and electricity rationing slowed growth to an average of approximately 5 percent between 2014 and 2017


```{r fig.width = 8, fig.height = 4, echo = FALSE}
plot(gdp_growth)
```

##### GDP Composition
* GDP is driven by services, comprising around 55% of GDP
* With the production of crude oil, industry has gradually expanded

```{r fig.width = 6, fig.height = 4, fig.asp = 0.5, echo = FALSE}
plot(gdp_shares)
```


##### Poverty in Ghana  
* Economic growth between 2012/13 and 2016/17 has __not been pro-poor__.  
* The middle class benefitted from growth, while those at the bottom end of
the distribution, especially the very poor, experienced a decline in consumption per adult
equivalent over the four-year period
* Poverty is largely a rural phenomenom
* Even though poverty in the Upper West Region is the highest among the 10 regions,
the region contributes less than 10 percent to national poverty because it is the
smallest region in Ghana in terms of population.

```{r fig.width=6, fig.height=8, fig.asp= 0.8, echo=FALSE}
plot(pov_plot)
```

##### Inequality
* Fundamental cause of rising poverty in the Volta and three Northern regions is lack of
economic growth in these regions
* Inequality is highest in the Upper West and Northern Regions
* In the Northern Region and Upper East Region, poverty increased because both the growth and redistribution deteriorated  

```{r fig.width=6, fig.height=8, fig.asp= 0.8, echo=FALSE}
plot(gini_plot)
```

##### Human Development
*Access to health services had declined over the years. The
proportion of the ill or injured who are likely not to consult a doctor or even visit a health
facility for treatment increased over the four-year period. 








Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

